Cartogram Mapping and Application to Cancer Data.

### 2019-06-24 review_article.Rmd
Editing text

Choropleth mapping methods effectively explore a relationship across a geographical space.
Thematic maps involve drawing administrative or other boundaries and filling the regions with colours, symbols or patterns to present levels, or rates, of the information communicated. @CIBMUK believes these 'representations are highly intuitive' as the familiar shapes help those looking at the maps to visually infer spatial relationships. 
The visualisations are true maps, used for visual inspection and can help to direct hypotheses about the spatial patterns across the physical space. 
These true maps rely on visual perception, a fallible metric as it can 'weight the shades from the largest areas more heavily.' The inferences derived from visual inspection, and the hypotheses developed can suffer from 'bias'. The use of thematic disease mapping was motivated by a 'desire to evaluate geographical patterns' @DMAHP. This technique has been seen as a 'trigger for questions' which drives research of spatial distributions. @MACM also recognise the intuition that can be derived from choropleths with the critique that they must be 'validated by rigorous statistical analyses.'

In epidemiology, choropleths are often used as a mapping tool in the study of the spatial distribution. 
Almost 100 years of cancer maps have explored the US and the UK with increased effectiveness in presentation of unbiased rates. 
The data collection methods, and administrative control within regions lends itself choropleth visualisation, for the presentation of cancer mortality rates across regions. Mortality rates are now presented as relative rates of risk across the population, an age adjusted to correct for the proportion of old people. Howe [-@HEDP] describes Stocks development of the standardised mortality ratios through the 1930s. The choropleth maps presented levels of cancer via hatchings on a black and white scale.

@DMAHP cites Cruickshank's (1947) discussion of using visuals as 'formal statistical assessment of the spatial pattern' as a major advancement.

@HEDP considered maps a research tool, dots maps use spots, choropleths and thematic maps shade areas and isopleths use isolines. These  techniques were commonly used before the advent of computer-assisted cartography.

@@

Cartograms are a thematic map, presenting statistical and geographical information like choropleths. The difference is the augmentation of the geographical areas.
This was originally implemented to allow a map base that represented the populations within regions. 
@HEDP argues that 'cancer occurs in people, not in geographical areas' and the map base should reflect this to avoid allocating 'undue prominence' to rural areas. It makes areas with greater populations bigger on the map space, but kept relationship with neighbours.
@DMAHP reduce the visual impact of large areas with small populations. The distortion of spatial areas is proportional to the denominator variable. This is implemented in an iso-demographic map, for approx equal population density.

Geographical and demographic base maps could be implemented after digital computer methods became available. Alternatively, the area on the map space can represent the value. There have been many algorithms presented, @SAIC investigate these implementations and the "statistical accuracy, geographical accuracy, and topological accuracy".

@@

<!-- Compress this to 2 sentence introduction about methods. Focus on which atlases have implemented this technique.--->
Cartograms keep spatial relationships of neighbours intact by keeping borders when adjusting sizes.
From a computer graphics perspective it is a problem of 'map deformation' to account for the value assigned to each area. @ACA discuss their implementation of three methods for creating value - by - area cartograms. @ECGC present the problem of maintaining shape presevation in non-contiguous cartograms.

The intention is to allow the map space to highlight the distribution of the variable. However drawing attention to large value areas and allowing reader to see all significant areas on a map, relies on them also knowing the difference between initial geography and new layout given by a cartogram.
Importance of the statistic in this area being communicated was more important than keeping geography recognisable.

eg. state by population

@@

We were able to create cartograms by access to data, access to tools, greater demand for insights.

@@

>>>>

Choropleth mapping methods effectively explore a relationship across a geographical space. 
Thematic maps involve drawing administrative or other boundaries and filling the regions with colours, symbols or patterns to present levels, or rates, of the information communicated. @CIBMUK believes these 'representations are highly intuitive' as the familiar shapes help those looking at the maps to visually infer spatial relationships. Using the topography as a base map is common for thematic maps of phenomena [@CD].
The visualisations are true maps, used for visual inspection and can help to direct hypotheses about the spatial patterns across the physical space. 
These true maps rely on visual perception, a fallible metric as it can cause viewers to 'weight the shades from the largest areas more heavily.' The inferences derived from visual inspection, and the hypotheses developed can suffer from 'bias'. The use of thematic disease mapping was motivated by a 'desire to evaluate geographical patterns' @DMAHP. This technique has been seen as a 'trigger for questions' which drives research of spatial distributions of diseases. @MACM also recognise the intuition that can be derived from choropleths with the critique that they must be 'validated by rigorous statistical analyses.'

In epidemiology, choropleths are often used as a mapping tool in the study of the spatial distribution. 
Almost 100 years of cancer maps have explored the US and the UK with increased effectiveness in presentation of unbiased rates. 
@HEDP considered maps a research tool, dots maps use spots, choropleths and thematic maps shade areas and isopleths use isolines. These  techniques were commonly used before the advent of computer-assisted cartography.
The data collection methods of cancer mortality rates across regions, and administrative control within regions lends itself choropleth visualisation. Mortality rates are now often presented as relative rates of risk across the population, and age adjusted to correct for the proportion of old people. Howe [-@HEDP] describes Stocks development of the standardised mortality ratios through the 1930s. The choropleth maps presented levels of cancer via hatchings on a black and white scale.

@DMAHP cites Cruickshank's (1947) discussion of using visuals as 'formal statistical assessment of the spatial pattern' as a major advancement.
@CD considers cartographic data analysis in practice. Presenting a clear guide for preparing cartograms for use with both qualitative and quantitative information.
The authors recognise that the creation of cartograms was largely the work of professional cartographers until the innovation of geographical information systems that welcomed 
map users as map creators. These systems are utilised depending on 'the effectiveness, efficiency, and satisfaction of the map products (Nielsen 1994)'.

Examples of atlases:
Kraak 1998, Kraak and Ormeling 2011
Bertin 1967

@@

Cartograms are a thematic map, presenting statistical and geographical information like choropleths. @AC suggests the difference is the intentional and desirable augmentation of the size, shape or distance of geographical areas.
This was originally implemented to allow a map base that represented the proportion of the population within each region. @AC considers this design a 'more socially just form of mapping' by giving all members of the population 'equitable representation'. This is known as an iso-demographic map, and the design inherently lends itself to epidemiolgy. @HEDP argues that 'cancer occurs in people, not in geographical areas' and the map base should reflect this to avoid allocating 'undue prominence' to rural areas. @DMAHP reduce the visual impact of large areas with small populations. The distortion of spatial areas is proportional to the denominator variable.

Alternatively, the area on the map space can represent the value. There have been many algorithms presented, @SAIC investigate these implementations and the "statistical accuracy, geographical accuracy, and topological accuracy".

@@

<!-- Intentionally preserve shape-->

@AC describes the process of creating a population cartogram for Canada, using the 1966 Census. An intentional goal of this project was to maintain contiguity, while attempting to keep the actual shape of places. The end result was a 'very accurate isodemo-graphic map of Canada'. This intentional design goal coincided with the rising interest in urban geography.

Cartograms keep spatial relationships of neighbours intact by preserving borders when adjusting sizes.
From a computer graphics perspective it is a problem of 'map deformation' to account for the value assigned to each area. @ACA discuss their implementation of three methods for creating value-by-area cartograms.

The intention is to allow the map space to highlight the distribution of the variable. However drawing attention to large value areas and allowing reader to see all significant areas on a map, relies on them also knowing the difference between initial geography and new layout given by a cartogram.

Requires knowledge of the methods

@@

@@

 @ECGC present the problem of maintaining shape presevation in non-contiguous cartograms.
